Why there are more species in several small patches vs few large patches:

### 2024-02-21 ms_SLOSS.Rmd
Updated ms

---
title: "Why there are more species in several small patches vs few large patches:
  a multi-species modeling approach"
bibliography: sloss_bib.bib
csl: "ecology-letters.csl"
output: pdf_document
---

@@

The SLOSS debate has been approached from several theoretical frameworks, but most of them were based on single-species models [@Fahrig2021]. Moreover, spatial dimension in several works has been neglected as mere noise around long-term equilibrium dynamics. [cita?] However, recent studies using spatially explicit simulations revealed that while high fragmentation leads to more local extinctions due to the reduced habitat size, at low levels of fragmentation, competitive exclusion is more prevalent, resulting in decreased diversity [@Rybicki2020].

@@

    where $d(x)$ is the probability that an individual disperses a distance $x$ from the parent. In all cases, we used $x_{min} = 1$. We use as parameter the mean $dispersal-distance$.

@@

For all simulations, we used a metacommunity of 100 species. The simulations start with the colonization of an empty habitat area. After 600 time steps fixed amount of 60% of habitat is subsequently subjected to destruction, employing regular or random block degradation to mimic habitat fragmentation due to habitat loss. <!-- why we don't use completely random degradation patterns -->

@@

## Results[e]

In our neutral model simulations with a migration rate of $0.0001$ and regular habitat patterns, species richness varied with habitat patch sizes, influenced by dispersal distance and the $\lambda$ parameter (Figure 1). Species richness increased with $\lambda$, with the highest richness at smaller patch sizes for higher $\lambda$ values. Conversely, a $\lambda$ of 3 or less showed reduced richness at the smallest patch sizes for lower dispersal distances. In these cases only migrants survive in the smallest patches, and no local populations are established (Figures S13 & S14).
The interplay of dispersal distances, habitat sizes, and $\lambda$ impacts community survival and species richness. For $\lambda < 4$ this suggests a threshold patch size influenced by these 3 parameters, for $\lambda = 4$ patch size decreases species richness. The differences between increasing $\lambda$'s almost disappear when dispersal distance is higher. Extended dispersal distances could enhance species survival in scenarios with low $\lambda$ and smaller patches but reduce richness beyond a certain threshold. Similar patterns are evident for Shannon diversity (Figure 2).

@@

For the simulations made with random block habitat patterns, the patches do not have the same sizes but there is a distribution with a minimum size determined by the patch-size parameter and a bigger maximum size (Figures S9 - S11). The pattern is similar to the regular case,  with a maximum richness for intermediate patch sizes and values of $\lambda$ higher than 2 the maximum is at the lower habitat patch size (Figure S1 & S2). The sizes of patches for the random block simulations are bigger so species richness and diversity are lower, except for the minimum patch size (Figure S7 & S8). Examples of simulations with random block and regular habitat patterns are shown in the supplementary material (Figures S9 & S32). In these figures we can observe that for the minimum patch size only migrants survive for regular habitat patterns with neutral and hierarchical models.

@@

Comparing the three versions of the model at $\lambda =2$, for higher dispersal the neutral model tend to have higher richness and Shannon diversity at higher patch sizes, followed by the hierarchical model and the neutral-with-habitat-selection model (Figure 3 & 4). At the lowest patch size, only in the neutral-with-habitat-selection have populations that can survive, because in the other model versions the individuals observed are migrants that cannot establish local populations and become extinct (See Figures S5 - S15). Thus an intermediate patch size with a maximum species richness is not observed for this model.

Examples of the final pattern of species and habitat distributions is shown in Figures S9- S32. Species spatial distributions closely match the spatial distribution previous to the habitat destruction for neutral and neutral-with-habitat-selection models. Is also clearly observed that for the hierarchical model there is a strong dominance of the more competitive species where Shannon Diversity is very low for higher dispersal rate (Figure 4 & S23).

@@

Edge effects are present in the neutral and hierarchical models, but not in the neutral-with-habitat-selection model. In the first two models, the edge effect is because individuals in the patch's border have fewer neighbors and the probability of extinction is higher.  In the neutral-with-habitat-selection model, the edge effect is not present because the individuals search for empty patches, so the individuals on the border of the patch have the same probability of extinction as the individuals in the interior of the patch. Thus when patches are smaller the edge proportion is higher and the edge effect is more important, but this plays a more important role in the minimal patch size for survival of species than for the species richness.

The spatial arrangement of habitat patches, contrary to some expectations, did not produce different species richness patterns regarding habitat patch size. The exceptions are stressed communities---represented in our models with lowest $\lambda$--- with low dispersal may have a continuous increment of richness with patch size, observed only in random block degradation patterns. This could be related to the optimal patch size for random block habitat loss patterns tends to be higher than for regular patterns, so we could hypothesize that the optimal was outside the simulated range for these stressed communities. The aggregation of patches in the random block pattern (implies aggregation of the degraded habitat) could be positive if individual habitat patches are very small because it allows some patches to surpass the minimum size required for survival. For bigger patch sizes it could be detrimental because the aggregation produces even bigger patch sizes that decrease species richness. A similar effect was observed on experimental landscapes [@Fletcher2023], aggregated patterns of habitat loss are detrimental to patch occupancy and eventually could produce the same effects observed in our models about richness.

Naturally, there are limitations to our modeling work. Firstly, it is important to note that more research is needed to apply our to determine the effect of fragmentation in real-world communities. We assume that all species have the same parameters, thus hypothetically determining these parameters for all the species, and using the mean to determine the responses to fragmentation would render an approximate value for the effects and thresholds. Also, an improved version of the model with different values of the parameters for each species could be developed and compared with the neutral model. The effects of fragmentation could be influenced by many other ecological/behavioural characteristics of the species not considered in the model, including environmental heterogeneity, matrix permeability, individual’s cost of dispersal, among others. Further, the size of the focal landscape compared to dispersal distances, which determines the scale of the dynamics, could influence quantitative results [@Rybicki2020]. Besides that, the approach of using mean values of the parameters of relatively simple multispecies models has been applied and tested to predict species biomass [@Fort2018].

@@

In terms of the cube hypothesis the between patch movement is influenced by the dispersal distance in relation to patch size, as the landscape size and the percentage of habitat destruction are constant, the between-patch distance is higher for bigger patch sizes except for the bigest size that is only one patch, also the number of patches is higher for smaller patch sizes thus between patch movement will be higher. Dispersal distance also increase between patch movement, but for higher dispersal distances the sizes of the patches are less important and the relationship tends to be SS = SL.   The spreading of risk is influenced by the $\lambda$, for smaller lambda spreading the risk would be more important, but to fully incorporate the spreading the risk effect the model should include disturbances. Very low dispersal distance favours still favours combined with low $\lambda$ favours SL > SS up to a threshold after wich the patch size is not important. The species spatial clumping is influenced by the dispersal distance, higher dispersal distance produce less clumpines but this is cofounded with patch movement, so the effect is difficult to disentagle. Thus the cube hypothesis is generally compatible with the results of our model.

>>>>

---
title: "Why there are more species in several small patches vs few large patches:
  a multi-species modeling approach"
bibliography: sloss_bib.bib
csl: "ecology-letters.csl"

output:
  pdf_document: 
    includes:
      in_header: margins.sty
---

@@

The SLOSS debate has been approached from several theoretical frameworks, but most of them were based on single-species models [@Fahrig2021]. <!--Moreover, spatial dimension in several works has been neglected as mere noise around long-term equilibrium dynamics [cita?]-->. However, recent studies using spatially explicit simulations revealed that while high fragmentation leads to more local extinctions due to the reduced habitat size, at low levels of fragmentation, competitive exclusion is more prevalent, resulting in decreased diversity [@Rybicki2020].

@@

    where $d(x)$ is the probability that an individual disperses a distance $x$ from the parent. In all cases, we used $x_{min} = 1$. We use as a parameter the mean $dispersal-distance$.

@@

For all simulations, we used a metacommunity of 100 species. The simulations start with the colonization of an empty habitat area. After 600 time steps, a fixed amount of 60% of habitat is subsequently subjected to destruction, employing regular or random block degradation to mimic habitat fragmentation due to habitat loss. <!-- why we don't use completely random degradation patterns -->

@@

## Results

In our neutral model simulations with a migration rate of $0.0001$ and regular habitat patterns, species richness varied with habitat patch sizes, influenced by dispersal distance and the $\lambda$ parameter (Figure 1). Species richness increased with $\lambda$, with the highest richness at smaller patch sizes for higher $\lambda$ values. Conversely, a $\lambda$ of 3 or less showed reduced richness at the smallest patch sizes for lower dispersal distances. In these cases, only migrants survive in the smallest patches, and no local populations are established (Figures S13 & S14).
The interplay of dispersal distances, habitat sizes, and $\lambda$ impacts community survival and species richness. For $\lambda < 4$ this suggests a threshold patch size influenced by these 3 parameters, for $\lambda = 4$ patch size decreases species richness. The differences between increasing $\lambda$'s almost disappear when dispersal distance is higher. Extended dispersal distances could enhance species survival in scenarios with low $\lambda$ and smaller patches but reduce richness beyond a certain threshold. Similar patterns are evident for Shannon diversity (Figure 2).

@@

For the simulations made with random block habitat patterns, the patches do not have the same sizes but there is a distribution with a minimum size determined by the patch-size parameter and a bigger maximum size (Figures S9 - S11). The pattern is similar to the regular case,  with a maximum richness for intermediate patch sizes and values of $\lambda$ higher than 2 the maximum is at the lower habitat patch size (Figure S1 & S2). The sizes of patches for the random block simulations are bigger so species richness and diversity are lower, except for the minimum patch size (Figure S7 & S8). Examples of simulations with random block and regular habitat patterns are shown in the supplementary material (Figures S9 & S32). In these figures, we can observe that for the minimum patch size only migrants survive for regular habitat patterns with neutral and hierarchical models.

@@

Comparing the three versions of the model at $\lambda =2$, for higher dispersal the neutral model tends to have higher richness and Shannon diversity at higher patch sizes, followed by the hierarchical model and the neutral-with-habitat-selection model (Figure 3 & 4). At the lowest patch size, only in the neutral-with-habitat-selection have populations that can survive, because in the other model versions the individuals observed are migrants that cannot establish local populations and become extinct (See Figures S5 - S15). Thus an intermediate patch size with a maximum species richness is not observed for this model.

Examples of the final pattern of species and habitat distributions are shown in Figures S9- S32. Species spatial distributions closely match the spatial distribution previous to the habitat destruction for neutral and neutral-with-habitat-selection models. Is also clearly observed that for the hierarchical model, there is a strong dominance of the more competitive species where Shannon Diversity is very low for higher dispersal rate (Figure 4 & S23).

@@

Edge effects are present in the neutral and hierarchical models, but not in the neutral-with-habitat-selection model. In the first two models, the edge effect is because individuals in the patch's border have fewer neighbors and the probability of extinction is higher.  In the neutral-with-habitat-selection model, the edge effect is not present because the individuals search for empty patches, so the individuals on the border of the patch have the same probability of extinction as the individuals in the interior of the patch. Thus when patches are smaller the edge proportion is higher and the edge effect is more important, but this plays a more important role in the minimal patch size for the survival of species than for the species richness.

The spatial arrangement of habitat patches, contrary to some expectations, did not produce different species richness patterns regarding habitat patch size. The exceptions are stressed communities---represented in our models with lowest $\lambda$--- with low dispersal may have a continuous increment of richness with patch size, observed only in random block degradation patterns. This could be related to the optimal patch size for random block habitat loss patterns, which tends to be higher than for regular patterns, so we could hypothesize that the optimal was outside the simulated range for these stressed communities. The aggregation of patches in the random block pattern (implies aggregation of the degraded habitat) could be positive if individual habitat patches are very small because it allows some patches to surpass the minimum size required for survival. For bigger patch sizes it could be detrimental because the aggregation produces even bigger patch sizes that decrease species richness. A similar effect was observed on experimental landscapes [@Fletcher2023], aggregated patterns of habitat loss are detrimental to patch occupancy and eventually could produce the same effects observed in our models about richness.

Naturally, there are limitations to our modeling work. Firstly, it is important to note that more research is needed to apply our to determine the effect of fragmentation in real-world communities. We assume that all species have the same parameters, thus hypothetically determining these parameters for all the species, and using the mean to determine the responses to fragmentation would render an approximate value for the effects and thresholds. Also, an improved version of the model with different values of the parameters for each species could be developed and compared with the neutral model. The effects of fragmentation could be influenced by many other ecological/behavioural characteristics of the species not considered in the model, including environmental heterogeneity, matrix permeability, and individual’s cost of dispersal, among others. Further, the size of the focal landscape compared to dispersal distances, which determines the scale of the dynamics, could influence quantitative results [@Rybicki2020]. Besides that, the approach of using mean values of the parameters of relatively simple multispecies models has been applied and tested to predict species biomass [@Fort2018].

@@

In terms of the cube hypothesis, the between patch movement is influenced by the dispersal distance concerning patch size, as the landscape size and the percentage of habitat destruction are constant, the between-patch distance is higher for bigger patch sizes except for the biggest size that is only one patch, also the number of patches is higher for smaller patch sizes thus between patch movement will be higher. Dispersal distance also increases between patch movements, but for higher dispersal distances the sizes of the patches are less important and the relationship tends to be SS = SL.   The spreading of risk is influenced by the $\lambda$, for smaller lambda spreading the risk would be more important, but to fully incorporate the spreading the risk effect the model should include disturbances. Very low dispersal distance favours still favours combined with low $\lambda$ favours SL > SS up to a threshold after which the patch size is not important. The species' spatial clumping is influenced by the dispersal distance, higher dispersal distance produces less clumpines but this is cofounded with patch movement, so the effect is difficult to disentangle. Thus the cube hypothesis is generally compatible with the results of our model.
